Model Based Analysis of Patient Derived Xenograft Growth Curves

### 2022-06-30 paper_Kuhlers.Rmd
final doc

---
title: "Model Based Analysis of Patient Derived Xenograft Growth Curves"
author: "Peyton Kuhlers"
date: "`r Sys.Date()`"
output:
  pdf_document:
    extra_dependencies: ["setspace", "float", "fancybox", "graphicx"]
bibliography: citations.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/american-medical-association.csl
---

@@

\doublespacing

@@

@@

Tumor volumes were analyzed using generalized linear mixed effects models.
Fixed effects included were time and the time-drug interaction.
The drug main effect was not included to model equal mean volume
at baseline amongst all treatment groups.
Volumes were assumed to be normally distributed and log-linearly
related to the covariates.
A random intercept and random time slope were included for each mouse.
Models were fit using the R package `lme4`/v1.1.27[@bates].
All confidence intervals are 95% Wald-type intervals.

@@

\begin{figure}[H]
\begin{center}
\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/growth_rescaled_days.pdf}}
\caption{Log tumor volumes over time.
Black lines indicate each mouses' growth curve.
Colored lines are simple linear regressions of log-volume versus time,
separated by each treatment group}
\end{center}
\end{figure}

The aim of the mixed effects model analysis was to model the trajectory
of the growth curve and determine if treatments slow tumor growth.
Therefore, we are interested in the growth rate (slope) of the tumor growth curve.
The mixed model coefficients represent the growth rates for each treatment,
relative to the control groups growth. The coefficients can
also be interpreted as the change in the ratio of the mean volume in the
treatment group compared to control per month.
Values less than 1 indicate the treatment reduces the growth rate relative to control.
Only gemcitabine and irinotecan were found to significantly reduce
tumor growth rates compared to control. The relative growth rates
were 0.60 (95% CI: 0.43, 0.85) and 0.71 (95% CI: 0.51, 0.99) which correspond to a
40% and 29% reduction in the growth rate relative to control, respectively (Fig. 2).

\begin{figure}[H]
\begin{center}
\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/mixed_forest.pdf}}
\caption{Forest plot of exponentiated mixed effects model interaction coefficients.
The coefficients represent the change in the ratio of mean volume in treatment versus control
over time. Red points indicate the growth rate is reduced in the treatment group,
while blue points indicate the growth rate is increased in the treatment group.
Horizontal bars indicate 95\% confidence intervals.}
\end{center}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/aft_models_overlaid_forest.pdf}}
\caption{Forest plot of AFT acceleration factors.
These represent the ratio of the mean time to 1500mm$^3$ in treatment versus control.
Red points indicate that time to 1500mm$^3$ is extended relative to control,
while blue points indicate the time is shortened.}
\end{center}
\end{figure}

@@

AFT models were also explored in this analysis. As aforementioned,
AFT models could be a complementary approach to area under the growth curve (AUC).
Both AUC and AFT analyses summarize the growth curve to a single number -- AUC or time to the endpoint
which can be useful with the growth curves have complex shapes.
For example, as seen in Fig. 1, some mice experience a period of tumor regression
followed by rapid tumor growth. Modeling this relationship is difficult
and may quickly become ad-hoc and produce overfitted models (especially
if the relationship is only observed in a handful of mice). An AUC or AFT analysis
can relatively easily handle this relationship since these mice will simply have
larger AUC and longer event times. The advantage of the AFT model is the results
are easily interpretable compared to AUC and can incorporate censoring for mice
that do not reach the volume endpoint.

The AFT models in this analysis largely reflected the mixed model results.
All drugs aside from vemurafenib extended the time to 1500mm^3^, with
carboplatin and erlotinib having the largest effect size.
While these results corresponds to the observed growth rates from the mixed model,
all of the acceleration factors were significant, while only a few of
the relative growth rates were. 
The discrepancy in effective drugs found by the two models could potentially be explained
by differences in baseline tumor volume. If two groups have the same slope
but different intercepts, the group with the higher intercept will reach
the endpoint first simply by being heavier at baseline, not because the treatment was effective.
While the experimental design and protocol justify the assumption,
its possible in practice inefficiencies in staging mice result in group level
differences at baseline. Additionally, the linear relationship may be poor approximations
of the true growth trajectory causing true differences in growth rates to be masked.
The discrepancy could also be simply
because small numerical differences in the growth rates (while statistically insignificant)
result in the overall time to 1500mm^3^ to be different.

Time to 1500mm^3^ is an artificial endpoint used for exploratory purposes in this analysis
and future experiments could set more meaningful volume endpoints that
may dramatically change these results.

>>>>

---
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    extra_dependencies: ["setspace", "float", "fancybox", "graphicx"]
mainfont: Times New Roman
fontsize: 12pt
bibliography: citations.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/american-medical-association.csl
---

@@

\newgeometry{top=2in,left=1in,right=1in}
\begin{titlepage}
\begin{center}

  MODEL BASED ANALYSIS OF PATIENT DERIVED XENOGRAFT GROWTH CURVES
  \vspace{1in}

  Peyton Kuhlers
  \vspace*{1in}

  A master's paper submitted to the faculty at the University of North Carolina at Chapel Hill
  in partial fulfillment of the requirements for the degree of Master of Science in the
  Department of Biostatistics in the Gillings School of Global Public Health.
  \vspace{1in}

  \begin{singlespace}
  Chapel Hill\\

  2022
  \end{singlespace}
  \vspace{1in}

\begin{doublespace}
  \begin{flushright}
    \begin{tabular}{l@{}}
      Approved by:\\Bahjat Qaqish\\Baiming Zou
    \end{tabular}
  \end{flushright}
\end{doublespace}

\end{center}
\end{titlepage}

\restoregeometry
\doublespacing

@@

\newpage

@@

Tumor volumes were analyzed using generalized linear mixed effects models.
Fixed effects included were time and the time-drug interaction.
The drug main effect was not included to model equal mean volume
at baseline amongst all treatment groups.
Volumes were assumed to be normally distributed with the log of the mean linearly
related to the covariates.
A random intercept and random time slope were included for each mouse.
Models were fit using the R package `lme4`/v1.1.27[@bates].
All confidence intervals are 95% Wald-type intervals.

@@

\begin{figure}[H]
\begin{center}
\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/growth_rescaled_days.pdf}}
\begin{singlespace}
\caption{Log tumor volumes over time.
Black lines indicate each mouses' growth curve.
Colored lines are simple linear regressions of log-volume versus time,
separated by each treatment group}
\end{singlespace}
\end{center}
\end{figure}

The aim of the mixed effects model analysis was to model the trajectory
of the growth curve and determine if treatments slow tumor growth.
Therefore, we are interested in the growth rate (slope) of the tumor growth curve.
The mixed model coefficients represent the growth rates for each treatment,
relative to the control group's growth. The coefficients can
also be interpreted as the change in the ratio of the mean volume in the
treatment group compared to control per month.
Values less than 1 indicate the treatment reduces the growth rate relative to control.
Only gemcitabine and irinotecan were found to significantly reduce
tumor growth rates compared to control. The relative growth rates
were 0.60 (95% CI: 0.43, 0.85) and 0.71 (95% CI: 0.51, 0.99) which correspond to a
40% and 29% reduction in the growth rate relative to control, respectively (Fig. 2).

\begin{figure}[H]
\begin{center}
\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/mixed_forest.pdf}}
\begin{singlespace}
\caption{Forest plot of exponentiated mixed effects model interaction coefficients.
The coefficients represent the change in the ratio of mean volume in treatment versus control
over time. Red points indicate the growth rate is reduced in the treatment group,
while blue points indicate the growth rate is increased in the treatment group.
Horizontal bars indicate 95\% confidence intervals.}
\end{singlespace}
\end{center}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/aft_models_overlaid_forest.pdf}}
\begin{singlespace}
\caption{Forest plot of AFT acceleration factors.
These represent the ratio of the mean time to 1500mm$^3$ in treatment versus control.
Blue points indicate estimates from the naive model
and red points indicate estimates from the imputed model.}
\end{singlespace}
\end{center}
\end{figure}

@@

AFT models were also explored in this analysis. As aforementioned,
AFT models could be a complementary approach to area under the growth curve (AUC).
Both AUC and AFT analyses summarize the growth curve to a single number -- AUC or time to the endpoint
which can be useful when growth curves have complex shapes.
For example, as seen in Fig. 1, some mice experience a period of tumor regression
followed by rapid tumor growth. Modeling this relationship is difficult
and may quickly become ad-hoc and produce overfitted models (especially
if the relationship is only observed in a handful of mice). An AUC or AFT analysis
can relatively easily handle this relationship since these mice will simply have
larger AUC and longer event times. The advantage of the AFT model is the results
are easily interpretable compared to AUC and can incorporate censoring for mice
that do not reach the volume endpoint.

The AFT models in this analysis largely reflected the mixed model results.
All drugs aside from vemurafenib extended the time to 1500mm^3^, with
carboplatin and erlotinib having the largest effect size.
While these results corresponds to the observed growth rates from the mixed model,
all of the acceleration factors were significant, while only a few of
the relative growth rates were. 
The discrepancy in effective drugs found by the two models could potentially be explained
by differences in baseline tumor volume. If two groups have the same growth rate (slope)
but different baseline volumes (intercepts), the group with the higher intercept will reach
the endpoint first simply by being heavier at baseline, not because the treatment was effective.
While the experimental design and protocol justify the equal baseline assumption,
its possible in practice inefficiencies in staging mice result in group level
differences at baseline. Additionally, the linear relationship may be poor approximations
of the true growth trajectory causing true differences in growth rates to be masked.
The discrepancy could also be simply
because small numerical differences in the growth rates (while statistically insignificant)
result in the overall time to 1500mm^3^ to be different.
Time to 1500mm^3^ is also an artificial endpoint used for exploratory purposes in this analysis
and future experiments could set more meaningful volume endpoints that
may change these results.
